\documentclass[phd]{infthesis}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[british]{babel}
\usepackage{microtype}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage[english=british,autopunct=false]{csquotes}
\usepackage[natbib=true,style=authoryear-comp,maxbibnames=6]{biblatex}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{wrapfig}
\usepackage{xfrac}
\usepackage{xspace}
\usepackage{mathcommon}
\usepackage[sc]{mathpazo}
\usepackage{hyperref}
\usepackage{expl3}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{mathpartir}
\usepackage{fancyvrb}
\usepackage{inconsolata}
\usepackage{multirow}

\frenchspacing

% Bibliography
\addbibresource{intro.bib}
\bibliography{intro}

% Text
\newcommand{\ie}{i.e.\xspace}
\newcommand{\eg}{e.g.\xspace}

% Referencing
\newcommand{\chp}[1]{\S\ref{chp:#1}}
\newcommand{\sct}[1]{\S\ref{sec:#1}}
\newcommand{\ssec}[1]{\S\ref{subsec:#1}}
\newcommand{\eqn}[1]{Eq.~\ref{eq:#1}}
\newcommand{\eqns}[2]{Eq. \ref{eq:#1} and \ref{eq:#2}}
\newcommand{\lem}[1]{Lemma~\ref{lemma:#1}}
\newcommand{\lems}[2]{Lemmas \ref{lemma:#1} and \ref{lemma:#2}}
\newcommand{\thm}[1]{Th.~\ref{thm:#1}}
\newcommand{\fig}[1]{Fig.~\ref{fig:#1}}
\newcommand{\diagram}[1]{diagram~\ref{eq:#1}}
\newcommand{\app}[1]{Appendix~\ref{app:#1}}
\newcommand{\mcite}[1]{\mrcolor{gray}{#1}} % missing cite
\newcommand{\defn}[1]{Def.~\ref{def:#1}}
\newcommand{\prop}[1]{Prop.~\ref{prop:#1}}

% Math
\renewcommand{\tuple}[1]{(#1)}
\DeclareMathOperator*{\expn}{exp}
\renewcommand*{\exp}[1]{e^{\,#1}} % \mathrm{e}^{#1}}
\renewcommand{\qedsymbol}{\ensuremath{\blacksquare}}
\newcommand{\partialto}{\rightharpoonup}
\newcommand{\id}{\vec{1}} % identity function
\newcommand{\mr}[1]{\mathrm{#1}}

\newcommand{\den}[1]{\llbracket #1 \rrbracket}
\newcommand{\m}[1]{\{\!| #1 |\!\}}
\newcommand{\M}[1]{\mathcal{#1}}
\newcommand{\MS}[0]{\mathrm{M}}
\newcommand{\SQ}[0]{\mathrm{S}}
\newcommand{\s}[1]{\underline{#1}}
\newcommand{\G}[0]{\Gamma}
\newcommand{\D}[0]{\Delta}
\newcommand{\mytt}{t\!t}
\newcommand{\myff}{f\!\!f}

\newcommand{\V}{\mathrm{V}}

\newcommand{\sel}{\mathrm{sel}}
\newcommand{\fold}{\mathrm{fold}}

\newcommand{\ms}{\mathrm{ms}}


\newtheorem{mydef}{Definition}
\def\dotminus{\mathbin{\ooalign{\hss\raise1ex\hbox{.}\hss\cr
  \mathsurround=0pt$-$}}}
\setlength{\tabcolsep}{8pt}
\renewcommand{\arraystretch}{1.0}

\newcommand{\match}{m}
\newcommand{\up}[1]{\uparrow\! #1}

\newcommand{\n}{\mathrm{n}}

% Other stuff
\newcommand{\maybe}[1]{\mrcolor{gray}{#1}}
\newcommand{\todo}[1]{\mrcolor{red}{TODO: #1}}

% rules
\newcommand{\ar}[2]{\mr{#1} \! = \! {#2}}

\setlength{\tabcolsep}{8pt}
\renewcommand{\arraystretch}{1.2}

\begin{document}
\chapter{Introduction}
% In light of this discussion I will also need to change the dicsussion in the
% next chapter

Life is produced and sustained through the organisation of entities at various
levels that interact to construct other autonomous entities. Organisation and
structure are therefore fundamental and defining properties of all living
systems. Some go as far as to say that (self-)organisation and the resulting
self-maintenance are the only common and defining characteristics of living
organisms and that they are primary to even natural selection in shaping the
forms of life that we see \citep[sometimes referred to as structuralism, for
example;][]{thompson1942growth, kauffman1992origins}. Even if we do not take such an
extreme view, it is still true that if we want to understand life we have to
understand the organisation that sustains it and the links between processes at
all scales defined by this organisation -- from genes to organisms and
ecosystems. Organisation here means both spatial organisation, for example cells
in a tissue, but also relations or interactions between processes that give rise
to other higher-level phenomena or between processes and the environment

% other less fundamental question could possibly be answered but at the core
% we need this
This view of organisation as a fundamental and defining principle of Biology
makes the subject distinct from other natural sciences, like Physics for example
and precludes a natural mode of enquiry, which is abstraction. The nature of the
questions is different. How can we abstract the details of an organism if those
details are exactly what makes the organism alive and in fact what we are trying
to understand? Since abstraction or at least complete abstraction is not
available, in order to answer biological questions we have to take a more
holistic and multi-scale view of the natural systems of interest.

%reductionist vs system (organisation)
Despite the importance of organisation in understanding life, historically the
focus in Biology has been on understanding individual mechanisms at the
molecular level. The reductionist view of life is that if cell is the main unit
of life then if we understand all the processes inside the cell we will
understand life as everything else follows.

There were, however, even if on the fringes or often completele outside
mainstream biology, few historical (and independent) threads of work that
recognised the importance of organisation and consequently of the systems
view. \citet{rashevsky_topology_1954} explained the need for organisation or
what he called the relational aspects of biology to complement usual approaches
that follow only the metric aspects of physical systems, which is the usual
abstraction in Physics where systems are viewed only by their quantifiable
properties (captured by numeric variables). \citet{rosen_relational_1958}
developed a theory, based on category theory, for the relational/organisational
aspects of biology and was a proponent of a systems view
\citep{rosen1991life}. \citet{varela_autopoiesis:_1974} developed the concept of
`autopoieisis' (self creation through organisation) as a necessary condition for
autonomous living entities. Following this, \citet{fontana_what_1994}
highlighted the importance of organisation as the definining characteristic of
living systems and also developed a theory of organisation as a basis for
biological understanding borrowing from computer science
\citep[$\lambda$-calculus in particular;][]{fontana_barrier_1996}. Finally, the
systems view entered the mainstream through the \emph{systems biology} movement
\citep{kitano2002systems} as a way to consolidate increasingly diverse
experimental molecular biology datasets.

% need for modelling
As the systems we are trying to understand become more complex, which is
particularly the case when we take the systems view, intuitive thinking and
mental models of the processes are not adequate tools to aid our
understanding. A useful tool in these cases is formal models where the systems
are represented using some formal languages based on a mathematical
intepretation, either directly or indirectly. An interesting view is that of
formal models as machines that turn the assumptions that we put in into them to
conclusions \citet{gunawardena_models_2014}. The unique advantage of formals
models compared to informal models (mental models) in Biology is that this
machine is guaranteed to be correct therefore providing a good tool for testing
our hypotheses (assumptions) just like an experiment would. Mental models
however do not provide this guarantee and therefore do not know if the fault is
in our assumption or in the animation of the model in our heads.

%also cite gunawardena2013biology
Historically, it is no surprise that formal models have been used in systems
models, for example for morphogenesis \citep{turing_chemical_1952} and neuronal
activity \citep{hodgkin_quantitative_1952\} although in some cases good
abstraction theories have been used in evolutionary genetics \citep[by
abstracting away the organism;][]{fisher} and in the small scale for chemical
kinetics \citep{michaelis-menten}, \citep[see
also;][]{gunawardena2013biology}. With the movement of systems biology as the
intellectual successor to the system and organisation theorists of the 20th
century we have seen an increase in mathematical models of cellular processes,
from metabolism, to the circadian clock, and .

% multi-scale models
% modelling with explicity representation of processes at multiple levels of
% organisation to understand the links between scales and how life or particular
% phenomena in life emerge through this, which cannot be explained only as a
% sum of their parts.
% also cite 
Taking the systems and modelling view to the extreme and remembering the
fundamental goal of understanding living organisms through their organisation we
can take a step further and expand our systems view from the cell to larger
scales, for example to the organism and beyond. These models that reach beyond
organisation on a single scale are usually called \emph{multi-scale models} and
they are more close to the historical spirit of understanding system
organisation to understand life, as we have seen above. For small
micro-organisms, comprehensive models link the metabolic and molecular level
with the cellular \citep{karr_whole-cell_2012} and population growth scales
\citep{weise_mechanistic_2015}. Work on more complex organisms has focused more
narrowly \citep[Virtual liver; virtual heart; bones]{} altough there are
ambitious collaborative project aiming at whole organisms \citep[virtual rat,
virtual human]{}.

%multi-scale models in plants
%plants
In plant science research is concentrated on the laboratory model species
\emph{Arabibopsis thaliana}, which offers an opportunity for broad understanding
that includes mechanistic models
\citep{chew_mathematical_2014, voss_modelling_2014}. 
%ecology and crops
Plants are interesting because modelling has been happening also on the other
side at the higher scale. There is ground to unify here
Unique challenges including languages
This further creates some problems social and technical, for example languages (\ref{})

Aims
In this thesis we are interested
This is where our work is placed buidling on this multiscale models (supra
multiscale) reaching to crop or ecology and evolution.


In the rest of this chapter I will give an overview of the uses of the
multi-scale plant models beyond the organism


\section{Multi-scale models -- from plants to ecosystems}
putting the organism back into evolution or ecology? That's the tagline?
introduce how and why questions
And then This will need some time to place the current work in other works in
plant ecology anad evolution

Nothings makes sense except in the light of population genetics

I think what needs justification is the need for putting the organism (the
individual in the population) There I can use the Doebeli etc.

\subsection{Ecology and evolution}
usually abstracting away the organism from this.
This has been useful in giving tractable representations and allowing general
principles to be uncovered. However, many details and the mechanistic foundation
of many of the assumptions is absent. A more fundamental view puts the organism
back into the equation and tries to explain the fitness differences starting
from 


\section{Multi-scale models -- from plants to crops}
This is perhaps easier? Need for multiscale, there are multiple papers
advocating for this, in fact there are whole movements

\subsection{Crop modelling}

\section{Languages for multi-scale biology}
why modelling? What are the purposes of models?
The fundamental difference in the question asked in Biology means that 
Only say a few general things, there's a whole chapter about this next.

\section{Contributions}
bullet points and references to later chapters
This need to be refutable, like a contract for the thesis

\printbibliography[heading=bibintoc]
\end{document}

% I think for the conclusion we somehow need to link to the multi-language for
% each scale and then one for the organisation, not sure