\documentclass[phd]{infthesis}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[british]{babel}
\usepackage{microtype}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage[english=british,autopunct=false]{csquotes}
\usepackage[natbib=true,style=authoryear-comp,maxbibnames=6]{biblatex}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{wrapfig}
\usepackage{xfrac}
\usepackage{xspace}
\usepackage{mathcommon}
\usepackage[sc]{mathpazo}
\usepackage{hyperref}
\usepackage{expl3}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{mathpartir}
\usepackage{fancyvrb}
\usepackage{inconsolata}
\usepackage{multirow}

\frenchspacing

% Bibliography
\addbibresource{thesis2.bib}
\bibliography{thesis2}

% Text
\newcommand{\ie}{i.e.\xspace}
\newcommand{\eg}{e.g.\xspace}

% Referencing
\newcommand{\chp}[1]{\S\ref{chp:#1}}
\newcommand{\sct}[1]{\S\ref{sec:#1}}
\newcommand{\ssec}[1]{\S\ref{subsec:#1}}
\newcommand{\eqn}[1]{Eq.~\ref{eq:#1}}
\newcommand{\eqns}[2]{Eq. \ref{eq:#1} and \ref{eq:#2}}
\newcommand{\lem}[1]{Lemma~\ref{lemma:#1}}
\newcommand{\lems}[2]{Lemmas \ref{lemma:#1} and \ref{lemma:#2}}
\newcommand{\thm}[1]{Th.~\ref{thm:#1}}
\newcommand{\fig}[1]{Fig.~\ref{fig:#1}}
\newcommand{\diagram}[1]{diagram~\ref{eq:#1}}
\newcommand{\app}[1]{Appendix~\ref{app:#1}}
\newcommand{\mcite}[1]{\textcolor{gray}{#1}} % missing cite
\newcommand{\defn}[1]{Def.~\ref{def:#1}}
\newcommand{\prop}[1]{Prop.~\ref{prop:#1}}

% Math
\renewcommand{\tuple}[1]{(#1)}
\DeclareMathOperator*{\expn}{exp}
\renewcommand*{\exp}[1]{e^{\,#1}} % \mathrm{e}^{#1}}
\renewcommand{\qedsymbol}{\ensuremath{\blacksquare}}
\newcommand{\partialto}{\rightharpoonup}
\newcommand{\id}{\vec{1}} % identity function
\newcommand{\mr}[1]{\mathrm{#1}}

\newcommand{\den}[1]{\llbracket #1 \rrbracket}
\newcommand{\m}[1]{\{\!| #1 |\!\}}
\newcommand{\M}[1]{\mathcal{#1}}
\newcommand{\MS}[0]{\mathrm{M}}
\newcommand{\SQ}[0]{\mathrm{S}}
\newcommand{\s}[1]{\underline{#1}}
\newcommand{\G}[0]{\Gamma}
\newcommand{\D}[0]{\Delta}
\newcommand{\mytt}{t\!t}
\newcommand{\myff}{f\!\!f}

\newcommand{\V}{\mathrm{V}}

\newcommand{\sel}{\mathrm{sel}}
\newcommand{\fold}{\mathrm{fold}}

\newcommand{\ms}{\mathrm{ms}}


\newtheorem{mydef}{Definition}
\def\dotminus{\mathbin{\ooalign{\hss\raise1ex\hbox{.}\hss\cr
  \mathsurround=0pt$-$}}}
\setlength{\tabcolsep}{8pt}
\renewcommand{\arraystretch}{1.0}

\newcommand{\match}{m}
\newcommand{\up}[1]{\uparrow\! #1}

\newcommand{\n}{\mathrm{n}}

% Other stuff
\newcommand{\maybe}[1]{\textcolor{gray}{#1}}
\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}

% rules
\newcommand{\ar}[2]{\mr{#1} \! = \! {#2}}

\setlength{\tabcolsep}{8pt}
\renewcommand{\arraystretch}{1.2}

\begin{document}
\chapter{Conclusions}
% summary
No complex system can be understood by extrapolation of the properties of its
elementary components. In order to understand a complex system, like an
organism, one needs to consider explanations at various levels of detail. This
is especially true in biological systems where entities (processes and so on)
and their organisation give rise to properties of processes at the next level of
detail and they in turn to the next until you get to the level of a living
entity, the organism. In order to understand the organism (and consequently
life) we therefore need to understand the links between the explanations at the
various levels of detail.

The importance of organisation between components has been stressed before and
it is in fact the fundamental concept of systems theory. While the focus in
Biology has been traditionally on single components the systems view has found
its way into mainstream biological thinking with the systems biology
movement. The stress on organisation and its constructive nature also leads to
questions about representation. The most widely used technical languages for the
representation of physical systems come from the dynamics world. Since the
questions there are usually different, abstracting a system to its properties
through variables and ignoring organisation is usually adequate. The
constructive organisation of biological systems though poses different
requirements although the value of abstraction still remains. We present
\emph{Chromar} a language that combine organisational aspects through the
explicit representation of discrete entities, agents, and non-organisational
aspects through abstract properties of these entities, attributes. The concept
of abstraction is there in extensions to the expression language used for the
dynamics of the attributes. Observables recognise the need for explanations at
different levels by allowing an explicit functional link between abstract
quantifiable properties and more explicit representations. Fluents, again, allow
abstraction via the description of system properties using time-dependent
deterministic value for describing parts of the system we do not wish to model
mechanistically.

In plant research the focus has always been on the model species
Arabidopsis. Functional-structural models have focused on the constructive
organisation of plant organs (development) as well as functional aspects like
metabolism. The Framework Model is inspired by these but adds more breadth by
explicitly including a genetic circuit and phenology models to time the
conceptual development towards flowering. The FM covers one part of plant
development but does not extend to reproduction, which means that it has limited
applicability in ecological studies. Ecological studies on the other hand
consider the full lifecycle but of an abstract version of plants where its
construction and physical development are absent. We presented \emph{FM-life}
an extension to the FM to the full lifecycle including reproduction and its
scale-up to the population level via a clustering approoach to tractable
simulate population of plants over multiple decades in different genotype x
environment scenarios.

Finally, we presented another more engineering oriented use of multi-scale plant
model as devices for optimising plant traits. Instead of the more traditional
genome engineering approach we focused on engineering the environment during
their development to achieve specific growth-related traits.

After this brief summary of the work presented in this thesis, we next go
through and dicuss each aspect of our work in the context we have set above to
understand its limitations, implications that follow, and possible future work
that it suggests.

\section{Chromar and representations of multi-scale biology}
Biology is modular. This includes "visibly" differentiated substructures like
the cell or organs, for example, given in spatial terms but also other entities
that form units in a non-spatial way that could be identified by the strength of
the interactions between their components or their functional goal. For example,
one might identify functional units in reaction networks
\citep{ederer_approach_2003} and hierarchical modules in metabolic networks
\citep{ravasz_hierarchical_2002}. These modules are organised either by
interaction -- more so the case in non-spatial modules -- or by physical forces
-- more the case in spatially distinguished modules like cells
\citep['Biological and Physical systems' section]{simon1962architecture}.

As we have pointed out, in order to understand complex biological systems, like
entire organisms, we need explanations at multiple levels of detail. At each
level of detail we can identify different modules, which can either be treated
as undivided wholes or as organisations of their elementary components/modules
\citep{varela_autopoiesis:_1974}. The links between the functioning of modules
at different levels of detail are crucial for fundamental questions in
Biology. We will call these aspects of biology that deal with modules, their
organisation (interaction), and hierarchy as the \emph{organisational} aspects
of systems. Importantly this organisation is constructive (dynamic).

Explicitly representing these links in organisational terms in detail, however,
is probably impractical in many cases. Even when considering the elementary
components of a module and their organisation, certain aspects of the
functioning of a module are still more practically captured through some
abstraction to pick out properties of interest that perhaps are important for
the links to the next level of detail.


Language question 



For the organisational parts of biology there is not much choice. We need
discrete objects relations and dynamics that add/create objects and
relate/unrelate objects.

Shortcomings of Chromar:
in organisational aspects
Most basic structure + dynamics
No explicit relation although they can be encoded
Connections

Inside the modules
Non-organisational aspects
Simple, again, but useful for a large range of application
Metric quantities and others

For the non-organisational parts the choice is greater since the
non-organisational parts are the abstractions that the modeller uses. Depending
on the system a different abstraction might be useful in understanding and
interpeting the process.

made the tie between non-organisational and organisational aspects
stronger. through fluents and observables
abstract vs non-abstract spaces
Make the variables more first-class

So that they represent whole models perhaps and their connection over multiple
level of explanation the abstract to the more concrete..enriched expressions
suggest a way that this could be done
could be combined

show how examples to see the challenges to advance their resolution
refer back to how things in the previous chapter lead to the realisation
discussed here
If I had more time + future

(mention database ideas that could mean borrowing ideas from databases for, say,
efficient implementation etc.)
+ Laurent paper
observables over a single agent
which suggests --->
--->

There are ways (see Ricardo's work to abstract a system that includes some form
of organisation and one level of detail to another.
Observables are not like that


\subsection{Multi-scale models that combine}
What does the Chromar experience suggest?
Organisational aspects through Chromar or an object-based lang to capture either
things that we consider objects or abstract objects like processes (see
Barabasi)
The non-organisational aspects can be captured in any relevant abstraction
depending on the level of understanding Can be even treated as undivided wholes
if needed

Practically will need either mappings between formalisms or more pragmatically
something like mois cis\_interface

mois cis etc. combination of models is not a model!!


\subsection{Practical considerations}
None of these are important if people cannot use these things
Given the above library of models with 



\section{Organism-centred evolutionary ecology}
effect of one mechanism on context? This theme is also important for the
previous section, abstract then mechanisms can be added later. It might be
interesting to see if we understand the effects of a single mechanism on
higher-levels of organisation.
Not always at the mechanism level -- see Marr's levels

Since we're plant biologist and the plant is where we have mechanims information
we would like to see the effect on ecology and a plant centred

With phenology we kind of understand at hte first level of understanding, going
to the second level of mechanism or of how this is realised we need the organism


Experimental validation?


\section{Multi-scale organism models for engineering}
Finally,
for these I'm not sure we need the organism. The organism is there for the
fundamental questions
We can engineer the environemnt etc.

Why do we use a multi-scale detail model?







\printbibliography[heading=bibintoc]
\end{document}